---
title: "RDS ICA"
author: "B242415"
date: "2023-12-03"
output:
  word_document: default
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Importing Data and Libraries

#### Required files: 
1. data_all.csv - .csv file containing sample names in row 1 and gene names in column 1.\
\
2. gene_annotation.csv  - .csv file containing gene names and their type and long names.\
\
3. genelist_72.csv - .csv files containing genes of interest to be included in analysis.\
\
4. sample_annotation.csv - .csv file containing sample names and their treatment groups.\

#### Required installed libraries: 
1. pheatmap


```{r import_data, echo=FALSE}
#import libraries
library(pheatmap)
#import data
try(data_all <- read.csv("data_all.csv"))
try(gene_annotation <- read.csv("gene_annotation.csv"))
try(genelist <- read.csv("genelist_72.txt"))
try(sample_annotation <- read.csv("sample_annotation.csv"))

#error checks for empty/missing data
if(exists("data_all") && exists("gene_annotation") && exists("genelist") && exists("sample_annotation")){
  print("Imported files!")
} else {
  print("At least one or more files has not been imported. Quitting...")
  quit()
}

```
\
\

## Data Filtering

Extracting data from data_all.csv with only genes listed in genelist_72.csv into data_filtered dataframe.\


```{r heatmap, echo=FALSE}
#1. Get only data from genes in genelist into data_filtered
data_filtered <- data.frame()

genelist_lengthbefore <- as.numeric(length(genelist[,1]))

genelist <- unique(genelist)

for(selectgene in genelist[,1]){
  for(i in 1:length(data_all[,1])){
    if(selectgene == data_all[i,1]){
      data_filtered <- rbind(data_filtered, data_all[i,])
    }
  }
}

#report duplicates
if(length(genelist[,1]) != genelist_lengthbefore){
  print(paste((genelist_lengthbefore - length(genelist[,1])), "duplicate(s) found in genelist.csv"))
}

print(paste(length(data_filtered[,1]) , " gene(s) seleced from data_all.csv for further processing.", sep=""))

#report missing genes that are in genelist but not in data_all
if(length(data_filtered[,1]) != length(genelist[,1])){
  print(paste( (length(genelist[,1])-length(data_filtered[,1])), "gene(s) in genelist is not in data_all.csv... Continuing without them."))
}

```
\
\

## Data Processing

log scale the data into logdata dataframe.

```{r data_processing, echo=FALSE}
#2. log scale the data 
logdata <- data_filtered

countna=0
for(val in 1:length(data_filtered[,1])){
  for(sample in 1:length(data_filtered[1,])){
    if(is.na(data_filtered[val,sample])){
      countna = countna + 1
    }
    
    logdata[val, sample] <- log(data_filtered[val, sample])
  }
}

#report na value counts
print(paste(countna,"NA values in data_filtered"))

##2a. renaming colnames of logdata df
i=1
for (gene in logdata$`log_X`){
  logdata$`log_X`[i] <- exp(gene)
  i = i+1
}

plotdata <- cbind(data_filtered, logdata[,2:length(logdata)])

print("Data log scaled.")

```
\
\

## Extracting annotations and long names
Extracting gene type and treatment type annotations from sample_annotation.csv and gene_annotation.csv.\
Also renaming gene names into their long name for heatmap plots.

```{r annotations, echo=FALSE}
#3. Annotate gene row with type of gene genetype
genetype_vector <- c()
for(gene in data_filtered$X){
  i=0
  for(annotation_gene in gene_annotation$X){
    i = i + 1
    if(gene == annotation_gene){
      genetype_vector <- c(genetype_vector, gene_annotation$Type[i])
      break
    }
  }
}

#check if all genes are in gene_annotation
if(length(genetype_vector) != length(data_filtered[,1])){
  print(paste( (length(data_filtered[,1])-length(genetype_vector)),"genes not included in gene_annotation.csv. Quitting... "))
  quit()
} else {
  print("All genes annotated with gene type")
}


genetype_df <- data.frame(genetype_vector)
rownames(genetype_df) <- rownames(data_filtered)

#4. Annotate each sample with treatment group
treatment_vector <- c()
for(sample in colnames(data_filtered)[colnames(data_filtered) != "X"]){
  i=0
  for(annotation_sample in sample_annotation$SampleName){
    i = i + 1
    if(sample == annotation_sample){
      treatment_vector <- c(treatment_vector, sample_annotation$TreatmentGroup[i])
      break
    }
  }
}

#check if all samples are in sample_annotation
if(length(treatment_vector) != length(data_filtered[,2:length(data_filtered[1,])])){
  print(paste( (length(data_filtered[2:length(data_filtered[1,]),])-length(treatment_vector)),"genes not included in gene_annotation.csv. Quitting... "))
  quit()
} else{
  print("All samples annotated with treatment groups")
}

treatment_df <- data.frame(treatment_vector)
rownames(treatment_df) <- colnames(data_filtered)[colnames(data_filtered) != "X"]

#5. Replace gene names with long names
longname_vector <- c()
for(gene in data_filtered$X){
  i=0
  for(annotation_gene in gene_annotation$X){
    i = i + 1
    if(gene == annotation_gene){
      longname_vector <- c(longname_vector, gene_annotation$LongName[i])
      break
    }
    
  }
}

#check if all longnames present
if(length(longname_vector) != length(data_filtered[,1])){
  print(paste( (length(data_filtered[,1])-length(longname_vector)),"gene long names not included in gene_annotation.csv. Quitting... "))
  quit()
} else {
  print("All genes renamed to their long names")
}

rownames(logdata) <- longname_vector
rownames(genetype_df) <- longname_vector


```
\
\

## Heatmaps

```{r heatmaps, echo=FALSE}
#6. Plot heatmaps
##6i. Rename annotation cols
colnames(genetype_df) <- "Gene Type"
colnames(treatment_df) <- "Treatment"
##6a. genes and samples clustered
pheatmap(logdata[,2:length(logdata)], 
         annotation_row = genetype_df, 
         annotation_col = treatment_df,
         treeheight_row = 25,
         treeheight_col = 25,
         angle_col=315,
         fontsize_row=7)
##6b. only genes clustered
pheatmap(logdata[,2:length(logdata)], 
         annotation_row = genetype_df, 
         annotation_col = treatment_df, 
         cluster_cols=FALSE,
         treeheight_row = 25,
         treeheight_col = 25,
         angle_col=315,
         fontsize_row=7)


```
